---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
##Load the data
data <- read.csv("activity//activity.csv", stringsAsFactors = FALSE, na.strings = "NA")
##Load the libraries we will use
library("lubridate")
library("sqldf")
```



## What is mean total number of steps taken per day?
```{r}
##Get the total number of steps for each date and plot it
stepsbyday <- sqldf("select sum(steps) as steps from data group by date")
hist(stepsbyday$steps)
```

```{r}
##Get the mean and the median
sqldf("select avg(steps) from stepsbyday")
median(stepsbyday$steps, na.rm = TRUE)
```


## What is the average daily activity pattern?
```{r}
##Get the average number of steps per interval and plot it
stepsbyinterval <- sqldf("select interval, avg(steps) as steps from data group by interval")
plot(stepsbyinterval$steps, stepsbyinterval$interval, type = "l", lab=c(5,24,4), las =1)
```
The average daily activity pattern shows significant activity from 8-9 with an average amount of activity across the rest of the general waking period that drops off near the end of the waking period.
```{r}
##Grab the max steps and figure out which interval it occured in
sqldf("select interval, steps from stepsbyinterval where 
	steps = (select max(steps) from stepsbyinterval)")
```

## Imputing missing values
```{r}
##Get all the records with missing steps
sqldf("select count(0) from data where steps is null")
##Check and see the distribution of the missing records across each interval 
sqldf("select count(0), interval from data where steps is null group by interval")
##Check and see the distribution of the missing records across each date
sqldf("select count(0), date from data where steps is null group by date")
##(note that there are 288 intervals per date in our data)
```
Missing data is uniformly distributed across each day and interval. The data seems to be missing each day completely and nothing else. We will take the average for each interval and add it for each day that it is missing
```{r}
fillData <- sqldf("select * from stepsbyinterval left join (select distinct date from data where steps is null) d")
datafill <- sqldf("select * from data where not steps is null")
datafill <- sqldf(c("insert into datafill(steps, date, interval) select steps, date, interval from fillData", "select * from datafill"))
stepsbyday <- sqldf("select sum(steps) as steps from datafill group by date")
hist(stepsbyday$steps)
sqldf("select avg(steps) from stepsbyday")
median(stepsbyday$steps, na.rm = TRUE)
```
From the plot, we can see that adding the missing data as averages causes the average occurance to increase sharply. Since the data is uniformly distributed across each day and time period simply removing the missing measurements does not affect our interval average or our daily average.



## Are there differences in activity patterns between weekdays and weekends?
```{r}
##get the data and transform it into two datasets, one for weekdays and one for weekends
##note we are using the filled in data so any differences between the two are less distinct
wdata <- transform(datafill, date = weekdays(ymd(date)))
wdata <- sqldf("select steps, case when date = 'Saturday' or date = 'Sunday' then 'Weekend' else 'Weekday' end as dayType, interval from wdata where not steps is null")
wdata <- sqldf("select avg(steps) as steps, dayType, interval from wdata group by dayType, interval")
weekend <- sqldf("select steps, interval from wdata where dayType = 'Weekend'")
weekday <- sqldf("select steps, interval from wdata where dayType = 'Weekday'")
##plot the data side by side
par(mfrow=c(1,2))
plot(weekday$steps, weekday$interval, type = "l", lab=c(5,24,4), las =1)
plot(weekend$steps, weekend$interval, type = "l", lab=c(5,24,4), las =1)
```
Both data sets show a strong spike around 0800 and 0900 and strong low from 2200 to 0600

The weekday data drops off during business hours from 0900 to 1800 with a small pickup afterwards



